Methods

### 2021-10-21 Manuscript/Manuscript_Methods.Rmd
added missing newlines and added section on terms relevance

### Text feature extraction
The collected citation data have a number of fields characterizing a scientific publication.
The framework models the relevance of a record based on the following fields:
title, abstract, authors, keywords, MESH terms [@lipscomb2000medical].
A series of Natural Language Processing (NLP) techniques [@baeza1999modern; @marshall2015systematic; @ananiadou2006text] are employed to transform the textual information in these fields into features for machine learning through a bag-of-words approach [@marshall2015systematic].
The processing of free text fields (title, abstract) includes: tokenization (i.e., extracting the terms), common stopwords (i.e. sentence components bringing no meaning) removal, part-of-speech filtering (only nouns, adjectives, verbs and untagged terms are retained), and lemmatization of the terms (i.e. reduction to their base grammar form).
Text processing for authors, keywords and MESH terms identifies logical units (e.g., author's full names, composite keywords) and extracts them.\
Terms appearing in less than 5% of the labelled documents are removed from negative records.
All terms in the positive set are kept to increase sensitivity at the cost of specificity.\
Some terms tend to co-appear in records (non-consecutive ngrams, nc-ngrams), often carrying a particular meaning when copresent.
To detect nc-ngrams, we generated a word network representation [@rousseau2015graph] posing edges between terms with a cosine similarity in terms of record co-occurrence > 0.5.
We extracted the maximal cliques in the network [@eppstein2010listing] representing highly correlated groups of terms;
These groups of terms are added to the data set as individual features.
Only nc-ngrams comprising a maximum of ten terms are kept.\
A second network is built using a co-occurrence threshold of 0.9.
In this case, the cliques represent terms that always appear together and therefore can be considered redundant (i.e. they do not need to be considered separately).
These terms are merged to increase computation efficiency and reduce overfitting.\
The output is a Document-Term Matrix (DTM), with $N_d$ rows representing the records ($D_i$), $N_t$ terms column for the $t_{field}$ terms (divided by record field) and ${0,1}$ values whether $t_{field} \in D_i$.
We also enriched the DTM with features referencing the number of terms in each field to help the model scale term importance based on the field length.

@@

@@

* We fit a partition tree [@rpart] between the DTM and 800 samples from the PPD;
if a term is present multiple times in the DTM (e.g. both title and abstract), they are counted just one, and field term count features are removed.
This step generates a list of rules composed by $AND$/$NOT$ "conditions" made of terms/authors/keywords/MESH tokens, which together identify a group of records.
* For each rule, negative conditions (i.e., $NOT$ statements) are added iteratively, starting from the most specific one, until no conditions are found that would not also remove positive records.
* The extended set of rules is sorted by positive-negative record difference in descending order.
The cumulative number of unique positive records is computed and used to group the rules.
Rules inside each group are ordered by specificity.
* The researcher is then asked to review the rule groups, selecting one or more rules (useful if they convey different meaning) from each, or edit them (in case too specific positive or negative conditions were included).
It is possible to exclude a group of rules altogether, especially those with the worse sensitivity/specificity ratio.
* The selected rules are joined together by $OR$ statements, defining a subset of records with a sensibly higher proportion of positive records than the original one.
* Redundant rules (i.e., rules whose positive records are already included in more specific ones) and conditions (i.e., conditions that once removed do not decrease the total number of positive or do not increase the negative records) are removed.
* Finally, the rules are re-elaborated in a query usable on the major scientific databases.

@@

To evaluate the hyperparameter effect of performance, we set up a "grid search" [@claesen2015hyperparameter; @yang2020hyperparameter] on a prelabelled subset made of the first 1200 records from the first session dataset. See Supplemental Material S1 for the combinations of hyperparameter used and other details on the method.\
The framework tested each hyperparameter combination until four CR iterations with no positive records were returned or the whole dataset got labelled.\
For each combination, a performance score was computed as the product of *Efficiency* (1 minus the ratio of records that required review over the total) and *Sensitivity* (number of positive records found over the total of positives).
We then identified homogeneous "performance clusters" of hyperparameter values and scores using a partition tree.
For the rest of our study, we chose the best cluster of parameter value combinations and, inside the cluster, the best combination according to Sensitivity and Efficiency in this order.

>>>>

### Text feature extraction
The collected citation data have a number of fields characterizing a scientific publication.
The framework models the relevance of a record based on the following fields:
title, abstract, authors, keywords, MESH terms [@lipscomb2000medical].
A series of Natural Language Processing (NLP) techniques [@baeza1999modern; @marshall2015systematic; @ananiadou2006text] are employed to transform the textual information in these fields into features for machine learning through a bag-of-words approach [@marshall2015systematic].
The processing of free text fields (title, abstract) includes: tokenization (i.e., extracting the terms), common stopwords (i.e., sentence components bringing no meaning) removal, part-of-speech filtering (only nouns, adjectives, verbs and untagged terms are retained), and lemmatization of the terms (i.e., reduction to their base grammar form).
Text processing for authors, keywords and MESH terms identifies logical units (e.g., author's full names, composite keywords) and extracts them.\
Terms appearing in less than 5% of the labelled documents are removed from negative records.
All terms in the positive set are kept to increase sensitivity at the cost of specificity.\
Some terms tend to co-appear in records (non-consecutive ngrams, nc-ngrams), often carrying a particular meaning when copresent.
To detect nc-ngrams, we generated a word network representation [@rousseau2015graph] posing edges between terms with a cosine similarity in terms of record co-occurrence > 0.5.
We extracted the maximal cliques in the network [@eppstein2010listing] representing highly correlated groups of terms;
These groups of terms are added to the data set as individual features.
Only nc-ngrams comprising a maximum of ten terms are kept.\
A second network is built using a co-occurrence threshold of 0.9.
In this case, the cliques represent terms that always appear together and therefore can be considered redundant (i.e., they do not need to be considered separately).
These terms are merged to increase computation efficiency and reduce overfitting.\
The output is a Document-Term Matrix (DTM), with $N_d$ rows representing the records ($D_i$), $N_t$ terms column for the $t_{field}$ terms (divided by record field) and ${0,1}$ values whether $t_{field} \in D_i$.
We also enriched the DTM with features referencing the number of terms in each field to help the model scale term importance based on the field length.

@@

### Relevant terms' extraction
BART model it is possible to extract the proportion of times a term was used in a posterior tree over the sum of total inclusions of all variables [@kapelner2013bartmachine], as a measure of feature importance for the model.
We extracted the terms, the part of the citation data they were used in, the average "inclusion rate" among the ensemble models (over 10,000 inclusions) and its ratio over the standard deviation of this inclusion (inclusion stability, IS).
For each term we ran a Poisson regression to get its linear association with a positive label and reported it as Relative Risk (RR) with the number of standard errors as significance index (Statistic);
the comparison between the inclusion rate in the BART models and the linear association allows to spot relevant non-linear effects (i.e, the feature is relevant only in association with others).
We reported in the Results only the first (in order of inclusion rate) 15 terms with IS > 1.5 while the first fifty terms regardless of inclusion stability are reported in Supplemental Material S2.

@@

* We fit a partition tree [@rpart] between the DTM and 800 samples from the PPD;
if a term is present multiple times in the DTM (e.g., both title and abstract), they are counted just one, and field term count features are removed.
This step generates a list of rules composed by $AND$/$NOT$ "conditions" made of terms/authors/keywords/MESH tokens, which together identify a group of records.
* For each rule, negative conditions (i.e., $NOT$ statements) are added iteratively, starting from the most specific one, until no conditions are found that would not also remove positive records.
* The extended set of rules is sorted by positive-negative record difference in descending order.
The cumulative number of unique positive records is computed and used to group the rules.
Rules inside each group are ordered by specificity.
* The researcher is then asked to review the rule groups, selecting one or more rules (useful if they convey different meaning) from each, or edit them (in case too specific positive or negative conditions were included).
It is possible to exclude a group of rules altogether, especially those with the worse sensitivity/specificity ratio.
* The selected rules are joined together by $OR$ statements, defining a subset of records with a sensibly higher proportion of positive records than the original one.
* Redundant rules (i.e., rules whose positive records are already included in more specific ones) and conditions (i.e., conditions that once removed do not decrease the total number of positive or do not increase the negative records) are removed.
* Finally, the rules are re-elaborated in a query usable on the major scientific databases.

@@

To evaluate the hyperparameter effect of performance, we set up a "grid search" [@claesen2015hyperparameter; @yang2020hyperparameter] on a prelabelled "validation set" made of the first 1200 records from the first session dataset.
See Supplemental Material S1 for the combinations of hyperparameter used and other details on the method.\
The framework tested each hyperparameter combination until four CR iterations with no positive records were returned or the whole dataset got labelled.\
For each combination, a performance score was computed as the product of "Efficiency" (1 minus the ratio of records that required review over the total) and "Sensitivity" (number of positive records found over the total of positives).
We then identified homogeneous "performance clusters" of hyperparameter values and scores using a partition tree.
For the rest of our study, we chose the best cluster of parameter value combinations and, inside the cluster, the best combination according to Sensitivity and Efficiency in this order.
